---
title: "Repro.Research_Project1"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Analysis of 'a personal activity monitoring device':

### Step One: load and pre-process the data

- Obtain data from [link in Project 1 instructions](https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip)
- save to working directory as 'activity.csv'
- load dataset into R and check its composition

```{r Load Dataset and packages}
activity<-read.csv('activity.csv',header=TRUE)
activity<-as.data.frame(activity)
str(activity)
```

```{r message=FALSE,warning=FALSE}
library(ggplot2)
library(dplyr)
library(gridExtra)

```
#### ...now pre-process the dataset

- remove NAs

```{r}
activity2<-na.omit(activity, cols=steps, invert=FALSE)
str(activity2)
```

- then assign variable names to the columns

```{r Create sum of steps per day and variable names for analysis}
date<-activity2$date
date<-as.Date(date)
steps<-activity2$steps
interval<-activity2$interval
```

### Step 2: create a histogram for the total number of steps taken each day

- As one tick-mark per day looks untidy, each bar here is one day, but the scale is one tick-mark per two weeks

```{r Step 2 Plot 1}
ggplot(activity2,aes(as.Date(date),steps))+stat_summary(fun.y = sum,geom = "bar",color="black",fill="blue")+scale_x_date(date_breaks="2 weeks")+theme_bw()+xlab("Date")+ylab("Number of Steps Taken")+ggtitle("Sum of Steps Taken Per Day")
```

### Step 3: Mean and median number of steps taken each day

- Calculate how many steps are taken each day ('daySteps')
- Use this to calculate the mean number of steps taken per day ('dayMean')
- Also use daySteps to calculate the median number of steps taken per day ('dayMedian')

```{r Step 3}
daySteps <- tapply(activity2$steps, activity2$date, sum)
dayMean <- mean(daySteps, na.rm = TRUE)
dayMedian <- median(daySteps, na.rm = TRUE)

dayMean
dayMedian
```
 - As we can see (above), the mean number of steps per day is 10766.19 and the median number of steps per day is 10765

### Step 4: Time-series plot of the average number of steps

```{r Step 4}

ggplot(activity2,aes(as.Date(date),steps))+stat_summary(fun.y = mean,geom = "line",color="black")+scale_x_date(date_breaks="2 weeks")+theme_bw()+ylab("Mean Number of Steps Taken")+ggtitle("Mean Number of Steps Taken Per Day")+xlab("Date")

```

### Step 5: The five-minute interval which, on average, contains the maximum number of steps

- For this section we will look in more detail at each five-minute interval recorded throughout each day rather than at the distribution of the total number of steps per day
- It is unclear whether this is to be the five-minute interval per day with the maximum number of steps, or that for the entire dataset
- So - we will calculate both of these options
- Option 1: the five-minute interval with the maximum number of steps across the entire dataset:
- Identify the maximum number of steps taken in any one five-minute interval

```{r max interval for dataset}
summary(steps)
```
- Then, subset the dataset to find the date and interval which contributed this maximum number of steps

```{r subsection activity}
maxInterval<- activity2[which(activity2$steps=='806'),]
maxInterval
```

- Therefore, the largest five-minute interval across the entire dataset is 806 steps taken at interval 615 on the 27th of November, 2012


- Option 2: the five-minute interval, after averaging, which contains the maximum number of steps:
- First, we want to see what the distribution looks like for average number of steps taken per five-minute time interval:

```{r average interval}
intervalSteps <- aggregate(steps~interval,data=activity,FUN=mean,na.action=na.omit)
intervalSteps$time <- intervalSteps$interval
h <- ggplot(intervalSteps, aes(time, steps))
h+geom_line(col="black")+ggtitle("Average steps taken per interval")+xlab("Interval Value")+ylab("Average Steps Taken")+theme_bw()

```

- So, the five-minute interval in question is between intervals 750 and 1000. We can subset this and look, but first lets shorten the x-axis of this graph to 'zoom in' on a narrower range of intervals

```{r average interval zoomed}
intervalSteps <- aggregate(steps~interval,data=activity,FUN=mean,na.action=na.omit)
intervalSteps$time <- intervalSteps$interval
h <- ggplot(intervalSteps, aes(time, steps))
h+geom_line(col="black")+ggtitle("Average Steps Taken per Interval")+xlab("Interval Value")+ylab("Average Steps Taken")+xlim(800,1000)+theme_bw()

```

- Therefore, we want to sub-set the average interval data to see which specific interval between intervals 800 and 850 has the greatest number of average steps

```{r subsection average interval}
averageInterval<- intervalSteps[which(intervalSteps$interval>='800' & intervalSteps$interval<='850'),]
averageInterval
```

- This tells us that the largest five-minute interval for the averaged data is interval number 835, with circa 206 steps

### Step 6: Describe and show a strategy for imputing missing data

- Let's check again what percentage of the activity dataset has missing values:

```{r missing values}
steps1<-activity$steps
mean(is.na(steps1))
```
- So we are missing about 13% of entries for steps and need to impute these values

- We will attempt this using the mean number of steps observed for each time interval and using these as replacements for the NAs for that interval

```{r impute missing data}

imp <- tbl_df(activity)
imp %>% filter(is.na(steps)) %>% summarize(missing_values = n())

activity$ImputedSteps <- ifelse(is.na(activity$steps), round(intervalSteps$steps[match(activity$interval, intervalSteps$interval)],0), activity$steps)

activity3 <- data.frame(steps=activity$ImputedSteps, interval=activity$interval, date=activity$date)

```
- This gives us a new dataset - activity3 - which contains the imputed values for the missing data

```{r}
head(activity3)
```

### Step 7: Histogram of the total number of steps taken each day after missing values are imputed

```{r Step 7 Plot 1}
ggplot(activity3,aes(as.Date(date),steps))+stat_summary(fun.y = sum,geom = "bar",color="black",fill="green4")+scale_x_date(date_breaks="2 weeks")+theme_bw()+xlab("Date")+ylab("Number of Steps Taken")+ggtitle("Sum of Steps Taken Per Day - Including Imputed Data")
```

- Let's compare the results of the original activity dataset against the new one with imputed values:

```{r}
p1<-ggplot(activity2,aes(as.Date(date),steps))+stat_summary(fun.y = sum,geom = "bar",color="black",fill="blue")+scale_x_date(date_breaks="2 weeks")+theme_bw()+xlab("Date")+ylab("Number of Steps Taken")+ggtitle("Steps Per Day - Original Dataset")

p2<-ggplot(activity3,aes(as.Date(date),steps))+stat_summary(fun.y = sum,geom = "bar",color="black",fill="green4")+scale_x_date(date_breaks="2 weeks")+theme_bw()+xlab("Date")+ylab("Number of Steps Taken")+ggtitle("Steps Per Day - Imputed Data")

grid.arrange(p1, p2, ncol = 2)
```

### Step 8: Panel plot comparing the average number of steps taken per 5-minute interval across weekdays and weekends

- First, we have to identify which dates are weekdays, and which are weekends, by correctly formatting the dates in the new imputed dataset 'activity3'

```{r}
activity3$Date <- as.Date(activity3$date, format = "%Y-%m-%d")
head(activity3)
```

- Then we can create a new variable 'weekday' based on these dates and the weekday() function

```{r}

activity3$weekday <- weekdays(activity3$Date)
activity3$Type <- ifelse(activity3$weekday=='Saturday' | activity3$weekday=='Sunday', 'weekend','weekday')
head(activity3)

```

- Now that we know which intervals and dates are from weekdays and which from weekends, we can seperate these into two datasets

```{r}
WDay<-activity3[activity3$Type == "weekday", ]
WEnd<-activity3[activity3$Type == "weekend", ]


```

- ... and then plot them out side-by-side

```{r}
a1<-ggplot(WDay,aes(interval,steps))+stat_summary(fun.y = sum,geom = "line",color="black")+theme_bw()+xlab("Interval")+ylab("Number of Steps Taken")+ggtitle("Average Steps - Weekday")+ylim(0,10500)

a2<-ggplot(WEnd,aes(interval,steps))+stat_summary(fun.y = sum,geom = "line",color="red")+theme_bw()+xlab("Interval")+ylab("Number of Steps Taken")+ggtitle("Average Steps - Weekend")+ylim(0,10500)

grid.arrange(a1, a2, nrow = 2)
```

